Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-09-02 manu-draft.qmd
starting discussion revision and additions

A total of 460 restoration projects were documented in Tampa Bay and its watershed between 1971 and 2019.  These projects were divided among habitat types that included estuarine (n = 228), freshwater (n = 53), uplands (n = 119), and a mix of all three (n = 60). A total of 1,978 ha have been restored, whereas 12,930 ha and 42.8 km (as shoreline or tributaries) were enhanced. Forty partners were responsible for these projects, although some were from departments within the same agency. Eighty-nine living shoreline projects, seawall enhancements, and oyster reef installations were documented, with a total linear length of 18.2 km.

@@

The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap). Native lands (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).  Similarly, restorable lands (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).  Understandably, most of the native and restorable lands occurred in undeveloped areas in the north and southwest. Existing conservation lands (existing conservation native, existing conservation restorable) totalled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totalled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).  Reservation areas in the coastal stratum (reservation native, reservation restorable) totalled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).

@@

Identifying short-term (2030) targets and long-term (2050) goals for the extent of native habitats in Tampa Bay was informed by the assessment of current extents, past trends, and relative effort for past restoration and enhancement projects. Table 4 shows the targets and goals identified through this analysis and the associated rationale. For example, the targets and goals are established based on the current extent and the restoration potential.  If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent plus the restoration opportunity.  Conversely, other habitats with no identified restoration opportunity, or current extents are sufficient (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and restoration is not recommended.  Implicit in the targets and the goals is recurring re-assessment over time to evaluate progress and adjust expectations as appropriate.

@@

Because of multiple stressors such as encroaching development and climate change, habitat protection and restoration priorities should be tempered and "reality tested" by what is actually possible today, and what is possible in the future. Many native and potentially restorable habitats are limited, and there will always be restrictions on the financial resources that can be dedicated to public conservation land acquisition and habitat restoration activities.

This replicable method for setting restoration targets and goals provides a systematic attempt to identify habitat protection and restoration targets that are based on what is actually achievable within those limitations. It focuses on existing opportunities for all habitat types, and what is realistically possible in the future, rather than attempting to mimic previous ecological conditions.

@@

When viewed as a whole, the most significant and meaningful trends in the TBEP habitats of interest over the periods of record examined include: 1) the 75 percent gain in seagrasses since 1988; 2) the slight gains in emergent tidal wetlands (10%) and freshwater wetlands (2%) since 1990; and 3) the 39% loss in native upland habitats since 1990.

The intertidal zone in Tampa Bay is currently experiencing dynamic change, driven by sea-level rise and climate change, whereby mangrove forests are outcompeting salt marshes and salt barrens for the available niche space. Without increasing the total area of the intertidal zone, restoring a greater coverage of mangroves would reduce the niche space available for salt marshes and salt barrens. This phenomenon has been observed throughout the Gulf of Mexico, and has been attributed to both climate change (e.g., fewer freeze events) and sea-level rise [@Comeaux12].

The observed gains in wetlands (Table 3b) are likely a reflection of: 1) the effectiveness of state and federal wetland regulatory programs; and 2) the cumulative gains resulting from, primarily, publicly-funded habitat restoration projects (Table 4). Minor gains in some emergent tidal wetlands (e.g., salt barrens) may also be a reflection of the landward expansion of the complex suite of these habitats associated with climate change and sea-level rise. Gains in vegetated non-forested freshwater wetlands are related to the clearing of forested wetlands followed by the creation of herbaceous mitigation areas and stormwater systems.

The decrease in native uplands (Table 3b) is the result of continued development in the Tampa Bay watershed, combined with the lack of regulatory protection of native uplands. Attaining the target and goal will require concerted restoration of native upland habitats on existing conservation lands, as well as new conservation lands to offset the continued loss of these habitats to development, and amendments to existing planning, zoning and land development policies or regulations. While federal and state regulations related to listed species management impart some protection to certain rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats, such as pine flatwoods, are left largely unprotected. Unless local governments in the Tampa Bay watershed improve local protections for native uplands, such as strengthening language within comprehensive plans and development ordinances, this trend will likely continue.

@@

Increases observed in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation. While restoration activities date to 1971, few projects were completed prior to 1990, and from 1990-2010, an annual mean of 68 ha of habitat was restored. Over the past decade (2010-2019), this rate of restoration project completion has increased to over 81 ha/yr. Assuming that funding levels remain in the same range as the past decade, this annual mean can be used to set reasonable limits on restoration potential and targets.

@@

Four major types of disturbed sites around the Tampa Bay coastline have been identified as priority estuarine habitat restoration sites by TBEP stakeholders over the past two decades, including dredged holes, filled and spoil disposal areas, abandoned aquaculture ponds; and coastal borrow pits and stormwater ponds. Substantial opportunities also exist for upland restoration on reclaimed mined lands within the watershed.

There is a general consensus among restoration practitioners and natural resource managers that habitat restoration and management is most cost-effective on publicly-owned conservation lands and provides long-term benefits. Furthermore, given current development trends in the Tampa Bay watershed, public acquisition of remaining critical lands (e.g., coastal uplands, river floodplain wetlands) is a high priority, and some restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships [@Holl01; @Benson18]. Therefore, varied approaches to leverage resources, including traditional grants, partner funding, and use of volunteers for habitat restoration are recommended to maximize the potential for successful target and goal achievement.

### Rolling easements, mitigation, and restoration consortium

@@

TBEP will form a public-private partnership to provide the framework for the development of a coordinated approach linking regulatory (compensatory mitigation) and resource management (publicly funded habitat enhancement, restoration, and establishment) programs in the watershed. The creation of the Habitat Management Consortium is expected to provide benefits such as optimizing and improving the cost-effectiveness of habitat protection, restoration and mitigation activities in the watershed.

>>>>

A total of 460 restoration projects were documented in Tampa Bay and its watershed between 1971 and 2019. These projects were divided among habitat types that included estuarine (n = 228), freshwater (n = 53), uplands (n = 119), and a mix of all three (n = 60). A total of 1,978 ha have been restored, whereas 12,930 ha and 42.8 km (as shoreline or tributaries) were enhanced. Forty partners were responsible for these projects, although some were from departments within the same agency. Eighty-nine living shoreline projects, seawall enhancements, and oyster reef installations were documented, with a total length of 18.2 km.  Although projects were documented for the whole period of record, few projects were completed prior to 1990.  From 1990 to 2010 and from 2010 to 2019, an annual mean of 68 ha/yr and 81 ha/yr of habitat was restored, respectively.  These means were used to define appropriate expectations for future restoration, described below.

@@

The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap). Native lands currently protected, proposed for protection, or in the coastal stratum (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW). Similarly, restorable lands currently protected, proposed for protection, or in the coastal stratum (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed). Understandably, most of the native and restorable lands occurred in undeveloped areas in the north and southwest. Existing conservation lands (existing conservation native, existing conservation restorable) totalled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totalled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed). Reservation areas in the coastal stratum (reservation native, reservation restorable) totalled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).

@@

Identifying short-term (2030) targets and long-term (2050) goals for the restoration extent of native habitats in Tampa Bay was informed by the assessment of current extents, past trends, and relative effort for past restoration and enhancement projects. Table 4 shows the targets and goals identified through this analysis and the associated rationale. For example, the targets and goals are established based on the current extent and the restoration potential. If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent plus the restoration opportunity. Conversely, other habitats with no identified restoration opportunity, or current extents are sufficient (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and restoration is not recommended. Implicit in the targets and the goals is recurring re-assessment over time to evaluate progress and adjust expectations as appropriate.

@@

Priorities for habitat restoration should be informed and validated by current assessments and what is possible to achieve in the future. These priorities are necessary to define given anticipated impacts of land development and climate change, while also considering competing societal interests for the use of the environment and the reality of limited resources for land acquisition and restoration. Our approach balances these tradeoffs by providing a systematic workflow to identify targets and goals that is informed by current extent, past trends, and realistic effort from past projects. Further, locations are identified where these targets and goals could be achieved based on existing opportunities for protected or restorable habitats, including areas anticipated be impacted by coastal stressors (i.e., sea level rise and land development). This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate historical baselines. Methods are also provided using open source tools [@Beck22] that 1) allow for the most current datasets to be synthesized to update progress on achieving targets and goals, and 2) can be used in other locations with similar needs for identifying restoration priorities.

@@

Identifying appropriate targets and goals would not have been possible without a detailed assessment of current extent and past trends of native habitats in Tampa Bay and its watershed. The most notable trends included 1) the increase of seagrasses by 75%, 2) the increase in emergent tidal wetlands (12%) and freshwater wetlands (24%), and 3) the loss in native uplands (38%).

Seagrass recovery in Tampa Bay is a well-known success story that demonstrated how public-private partnerships can effectively reduce total nitrogen loads into Tampa Bay [@Greening14; @Sherwood17] . The nutrient reductions, primarily from point-source controls and advanced wastewater treatment, contributed to improvements in water quality and light environments that were favorable for seagrass growth. However, the most recent coverage estimate in 2020 showed a seagrass loss of 18% since the peak in 2016, falling below the target defined herein. These data were unavailable at the time the habitat restoration workflow was initially developed and recent seagrass changes have prompted bay managers to assess barriers in achieving the seagrass restoration goal. In particular, much of the seagrass losses have occurred in Old Tampa Bay, where recurring algal blooms of *Pyrodinium bahamense* have contributed to water quality decline [@Lopez19] . Ongoing research to understand mechanisms for mitigating these blooms will be critical for restoring seagrass in Tampa Bay.

Emergent tidal and freshwater wetlands in Tampa Bay have also experienced dramatic changes over the last three decades. Dual pressures from sea-level rise and climate change have impacted tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space. Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12]. As such, the identified targets and goals for mangroves indicate protection of mangrove habitats, without the need for additional restoration. However, mangroves will continue to expand in the intertidal zone contributing to additional losses of salt marshes and salt barrens. The reservation areas identified in @fig-oppmap represent critical remaining areas in the intertidal zone that could be protected to prevent additional losses of tidal wetlands. Likewise, gains in non-forested freshwater wetlands are a reflection of 1) the effectiveness of state and federal wetland regulatory programs, and 2) the cumulative gains from publicly-funded habitat restoration projects. A similar approach could be used for salt marshes or salt barrens.

The decrease in native uplands is the result of continued development in the Tampa Bay watershed (@fig-chgdatalluv) and lack of regulatory protection of these habitats. Attaining the target and goal will require restoration of upland habitats on existing conservation lands (i.e., restoration potential in @fig-restmap) and new conservation lands to offset the continued loss of these habitats to development.  The long-term conversion of restorable lands to developed (@fig-chgdatalluv) presents additional challenges for restoration of native uplands.  Amendments to existing planning, zoning and land development policies or regulations will be needed to address these issues. While federal and state regulations related to listed species management provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods).  Reductions in native uplands will likely continue unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development ordinances.

@@

There is a general consensus among restoration practitioners and natural resource managers that habitat restoration and management is most cost-effective on publicly-owned conservation lands and provides long-term benefits. Furthermore, given current development trends in the Tampa Bay watershed, public acquisition of remaining critical lands (e.g., coastal uplands, river floodplain wetlands) is a high priority, and some restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships [@Holl01; @Benson18]. Therefore, varied approaches to leverage resources, including traditional grants, partner funding, and use of volunteers for habitat restoration are recommended to maximize the potential for successful target and goal achievement. Increases observed in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation.  Assuming that funding levels remain in the same range as the past decade, estimates of past restoration effort can be used to set reasonable limits on restoration potential and targets.

Four major types of disturbed sites around the Tampa Bay coastline have been identified as priority estuarine habitat restoration sites by stakeholders over the past two decades, including dredged holes, filled and spoil disposal areas, abandoned aquaculture ponds; and coastal borrow pits and stormwater ponds. Substantial opportunities also exist for upland restoration on reclaimed mined lands within the watershed.

@@

### Rolling easements and mitigation

@@

### Limitations of the approach

Identifying restoration priorities was data intensive and would not have been possible without the resources available for the region. The workflow for identifying priorities required detailed and spatially-explicit datasets specific to the Tampa Bay watershed. Long-term datasets describing land use and cover in the watershed and the extent of in-bay habitats were necessary to categorize current extent and past trends. Similarly, supporting datasets included those that described existing and proposed conservation areas, soils, past restoration activities, and relevant spatial boundaries (i.e., watershed and coastline). Many of these datasets are available outside of Tampa Bay, although temporal and spatial resolutions may limit application to other areas. Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types. For example, inventories of hard bottom habitats, living shorelines, tidal tributaries, and artificial reefs were created through special studies or were available only as current estimates from regional entities. Tracking progress towards targets in goals for these habitats will require regular updates to these datasets that are not part of routine land use and cover maps.

An additional assumption of the workflow, particularly for tracking progress, is that projects reported by partners will ultimately manifest in map products. Specifically, restoration effort by regional partners is catalogued in the available restoration database, which not only depends on voluntary reporting, but also represents a source of information on restoration extent that is separate from land cover maps. An expectation is that acreage restored for a given habitat as reported by a partner will ultimately be shown as a change in land use and cover on regional maps. The temporal lag between an actual project and how it may be reflected in a GIS product is unknown, which may create a disconnect between the updates in achieving targets and goals as new data layers are released and the effort reported by partners. The spatial resolution of mapping products may also be insufficient to detect habitat changes as reported in the restoration database. For these reasons, projects reported by partners are currently summarized separately from the assessments above that depend on GIS layers. Additional work is needed to reconcile these datasets for more streamline reporting.
